---
title: "Suicide Data Analysis Project"
author: "Stats Stars"
date: "04/29/19"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

### Introduction

### Load Data and Packages & Organize Data

```{r load-data, message=FALSE}
library(tidyverse)
library(broom)
suicides <- read_csv("/cloud/project/data/master.csv")
```

```{r suicides}
suicides <- suicides %>%
  mutate(Year = year,
         Age = age,
         Sex = sex,
         Country = country,
         Generation = generation,
         Population = population,
         suicides_per_100k_pop = `suicides/100k pop`, 
         gdp_for_year = `gdp_for_year ($)`, 
         gdp_per_capita = `gdp_per_capita ($)`,
         HDI_for_year = `HDI for year`,
         country_year = `country-year`) %>%
  filter(year != "2016")
```

```{r relevel-generation-age}
suicides <- suicides %>%
  mutate(Generation = fct_relevel(Generation, "G.I. Generation", "Silent", "Boomers", "Generation X", "Millenials", "Generation Z"))
suicides <- suicides %>%
  mutate(Age = fct_relevel(Age, "75+ years", "55-74 years", "35-54 years", "25-34 years", "15-24 years", "5-14 years"))
```


### Suicides and Location

```{r country-plot, fig.height=12, fig.width=8}
topcountries <- suicides %>%
  group_by(country) %>%
  summarise(n = mean(suicides_per_100k_pop)) %>%
  arrange(desc(n))

ggplot(data = topcountries, mapping = aes(x = reorder(country,n), y = n)) +
  geom_bar(aes(y = n), stat = "Identity", fill="blue") +
  coord_flip() +
  labs(title = "Suicides per 100k people",
       subtitle = "by country",
       x = "Country",
       y = "Count")
```

```{r highest-rate-by-country}
topcountries %>%
 slice(1:10)

topcountries %>%
  filter(country == "United States")

topcountries %>%
  summarise(min = min(n), max = max(n),
            mean = mean(n), med = median(n),
            sd = sd(n), 
            q1 = quantile(n, probs = 0.25),
            q3 = quantile(n, probs = 0.75),
            iqr = (q3 - q1))
```

By sorting the data and differentiating the suicide rates of countries, we are able to look at and compare individual countries. For instance, lots of countries claim that they are the happiest nation. We can see if there is any comparison between happiness and rates of suicide in specific nations. For instnace, we decided to compare The Netherlands and Ukraine- Netherlands which is ranked #5 in the World Happiness Report while Ukraine is ranked #133. 


```{r extreem}
extreem <- suicides %>%
  filter(country %in% c("Netherlands", "Ukraine"))
```


```{r suicide-happiness, message=FALSE}
ggplot(data = extreem, mapping=aes(x = year , y = suicides_per_100k_pop, color = country)) + 
  geom_smooth()  +  
  labs(title = "Netherlands & Ukraine Suicides by Year",
       x = "Year", y = "Number of Suicides/100K Population")
```

As demonstrated in the visualization, there is a significant difference in the suicide rates per 100k people between people in The Netherlands and Ukraine. Although correlation does not indicate causation, it is curious that there seems to be a relationship between happiness/saddness and suicide rates in a nation.

### Suicides and Economic Situation over all Time

We wanted to see if there was a correlation between proportion of suicides and the economic situation in each country. The measure of the economic situation in this data set is GDP per capita measured in US$. We thus created a linear model that visualized the number of suicides per 100k of the population versus the GDP per capita. The graph used data from all countries in the dataset and from the years 1980-2015. There are a lot of data points on the graph and they are all quite scattered. When we fit a linear regression to the graph it gave the equation given below the graph and had an r-squared value also shown below the graph. The very low r-squared value and the very weak positive relationship the regression model gave, indicate that there is little, if any, relationship between GDP per capita and the number of suicides per 100k of the population.  

```{r create-dataframe-grouping-data-countryyear}
#aggregated all the suicides/100k pop data per country-year
suicides_grouped <- aggregate(suicides_per_100k_pop ~ country_year, data = suicides, FUN=sum)

#merged data set created with old one to have access to gdp data
suicide_economics <- merge(suicides, suicides_grouped, by.x = "country_year", by.y = "country_year") %>%
  select(country_year, suicides_per_100k_pop.y, gdp_per_capita)
```

```{r vis-suicide-gdp-per-capita}
ggplot(data = suicide_economics, mapping = aes(x = gdp_per_capita, y = suicides_per_100k_pop.y)) + 
  geom_point(size = 0.1, color = "blue") +
  geom_smooth(method = "lm", se = FALSE,
              col = "orange") +
  labs(title = "Suicides per 100k of the population vs. GDP per capita",
       x = "GDP per capita ($)",
       y = "Suicides per 100k of the population")
```

```{r linear-model-suicide-economics}
model_gdp_suicides <- lm(suicides_per_100k_pop.y ~ gdp_per_capita, data = suicide_economics)
tidy(model_gdp_suicides)
glance(model_gdp_suicides)$r.squared
```

Equation for the linear model is: (Suicides per 100k of the population-hat) = (153) + (0.0000243*GDP per capita($)) 

R-squared value is: 1.534907e-05

The intercept of this model is 153. This means that for a country with a $0 GDP per capita the number of suicides per 100k of the population would be expected to be, on average, 153.

The slope of this model is 0.0000243. This means that for every $1 increase in GDP per capita, the number of suicides per 100k of the population is expected to increase, on average, by 0.0000243.

The r-squared value is 1.534907e-05. This means that the GDP per capita in US$ explains 0.00153% of variability in the number of suicides per 100k of the population.

### Suicides and Economic Situation in 2015

Because the previous graph showed so much variability in how GDP per capita affected the proportion of suicides, we were curious to see if the relationship would prove to be clearer by focusing on a single time period. We also thought that GDP per capita might be an imperfect statistic in tracking relative economic affluence over a large period of time. For this reason we made the same visualization and linear regression model, but this time only using data from 2015. This model showed a slightly larger positive correlation between increasing GDP per capita ($) and number of suicides per 100k of the population, but the r-squared value was still not high enough to show any significant correlation. 
```{r create-dataframe-grouping-data-countryyear-2015}
suicides_current <- suicides %>%
  filter(Year == "2015")

#aggregated all the suicides/100k pop data per country-year
suicides_grouped_current <- aggregate(suicides_per_100k_pop ~ country_year, data = suicides_current, FUN=sum)

#merged data set created with old one to have access to gdp data
suicide_economics_current <- merge(suicides, suicides_grouped_current, by.x = "country_year", by.y = "country_year") %>%
  select(country_year, suicides_per_100k_pop.y, gdp_per_capita)
```

```{r vis-suicide-gdp-per-capita-2015}
ggplot(data = suicide_economics_current, mapping = aes(x = gdp_per_capita, y = suicides_per_100k_pop.y)) + 
  geom_point(size = 0.5, color = "blue") +
  geom_smooth(method = "lm", se = FALSE,
              col = "orange") +
  labs(title = "Suicides per 100k of the population vs. GDP per capita",
       subtitle = "in 2015",
       x = "GDP per capita ($)",
       y = "Suicides per 100k of the population")
```

```{r linear-model-gdp-2015}
model_gdp_suicides_current <- lm(suicides_per_100k_pop.y ~ gdp_per_capita, data = suicide_economics_current)
tidy(model_gdp_suicides_current)
glance(model_gdp_suicides_current)$r.squared
```
Equation for the linear model is: (Suicides per 100k of the population-hat) = (127) + (2.168189e-04*GDP per capita($)) 

The R-squared value is: 0.003774998

The intercept of this model is 127. This means that for a country in 2015 with a $0 GDP per capita the number of suicides per 100k of the population would be expected to be, on average, 127.

The slope of this model is 2.168189e-04. This means that for every $1 increase in a country's GDP per capita in 2015, the number of suicides per 100k of the population is expected to increase, on average, by 2.168189e-04.

The r-squared value is 0.003774998. This means that the GDP per capita ($) explains 0.377% of variability in the number of suicides per 100k of the population in 2015.

### Suicides and Generation

We wanted to also explore the link between generation and suicide rates. The boxplot shown below clearly shows that each older generation has higher suicide rates than the one after it, meaning that suicide rates are largely decreasing over time. This trend is mirrored in the trends of suicide rate over the years. 
There does seem to be a cluster of outliers in the Silent generation that is even higher than most of the outliers in G.I. Generation. When we did some further digging into the possible causes of this, we found that those abnormally high suicide rates among the Silent generation were almost all due to Korea's high suicide rates in 75+ year old men between the years of 2003 and 2015. When we did further research we found that the cause of those abnormally high suicide rates is because the elderly have no source of income because the country's pension system only began in 1988. As the country advances many young people leave their parents behind. And even if the young continue to financially support their parents, the older generation does not want to be a financial burden. This is causing their abnormally high suicide rates.

```{r vis-generation-suicides}
ggplot(data = suicides, mapping = aes(x = Generation, y = suicides_per_100k_pop)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "Suicides per 100k of the population vs. Generation",
       x = "Generation",
       y = "Suicides per 100k of the population")
```

### Suicides and Age Group

Next, we decided to look at what age people are most likely to commit suicide. The boxplot below shows a clear positive trend in increasing age and proportion of suicides. From the summary statistics shown below the graph, we can infer that a person 75+ years of age is almost two times more likely to commit suicide than someone between 25 and 34 years old. 
```{r vis-age-suicides}
ggplot(data = suicides, mapping = aes(x = Age, y = suicides_per_100k_pop)) +
  coord_flip() +
  geom_boxplot() +
  labs(title = "Suicides per 100k of the population vs. Age",
       x = "Age",
       y = "Suicides per 100k of the population")
```

```{r age-suicides-summarystats}
suicides %>%
  group_by(Age) %>%
  summarise(mean = mean(suicides_per_100k_pop),
            median = median(suicides_per_100k_pop),
            min = min(suicides_per_100k_pop),
            max = max(suicides_per_100k_pop),
            sd = sd(suicides_per_100k_pop),
            q1 = quantile(suicides_per_100k_pop, probs = 0.25),
            q3 = quantile(suicides_per_100k_pop, probs = 0.75),
            iqr = (q3-q1))
```


### Suicides and Year

Group number of suicides together by year to find number of suicides for each year:

```{r year-suicides}
year_suicides<- suicides %>%
  group_by(Year) %>% 
  summarize(n = sum(suicides_per_100k_pop))
```

Create visualization to represent the number of suicides for each year:

```{r year-suicides-visualization}
ggplot(data = year_suicides, mapping=aes(x = Year, y =  n)) + 
  geom_line()  +  
  labs(title = "Number of Suicides/100K Population for Each Year", 
       x = "Year", y = "Number of Suicides/100K Population")
```

### Suicides and Sex

Group number of suicides/100k population together by sex to find number of suicides/100k population for each sex:

```{r sex-suicide}
suicides_sex <- suicides %>%
  group_by(Sex) %>%
  summarize(n = sum(suicides_per_100k_pop))
```

Create visualization to represent the number of suicides/100k population for each year:

```{r sex-suicides-visualization}
ggplot(data = suicides_sex, mapping=aes(x = Sex, y =  n)) + 
  geom_col() +
  labs(title = "Number of Suicides/100K Population by Sex", 
       x = "Sex", y = "Number of Suicides per 100k Population")
```

Compare number of suicides/100k population for each sex per year:

```{r year-sex-suicides}
year_sex <- suicides %>%
  group_by(Year, Sex) %>% 
  summarize( n = sum(suicides_per_100k_pop))
year_sex
```

Create visualization to represent proportions of gender for number of suicides/100K population by year:

```{r line}
ggplot(year_sex, aes(x = Year, y = n, color = Sex)) +
    geom_line() +
    labs(title = "Changes in Male and Female Suicide Rates Over Time",
         x = "Year", y = "Number of Suicides per 100k People")
```


### Linear Modeling 

To figure out which variables influence suicides the most, we will be doing a linear regression. 

```{r full-model}
full_model <- lm(suicides_per_100k_pop ~ Age + Year + Country + Sex +
                   Generation + Population + gdp_for_year +
                   gdp_per_capita + HDI_for_year + country_year, data = suicides)
```

 
```{r select-model}
#selected_model <- step(full_model, direction = "backward")
```


### Shiny App

```{r shinyapp}
ui <- fluidPage(
    
    # Sidebar layout with a input and output definitions
    sidebarLayout(
        
        # Inputs: Select variables to plot
        sidebarPanel(
            
           # Select country
            selectInput(inputId = "country_data",
                      label = "Country",
                      choices = suicides$country,
                      selected = "`Albania`"),
           
            # Select variable for x-axis
            selectInput(inputId = "x", 
                        label = "X-axis:",
                        choices = c("Year", "Sex", "Age", "Generation", "Population", "gdp_for_year",
                                    "gdp_per_capita", "HDI_for_year"),
                        selected = "Year"),
            
            # Select variable for y-axis
            selectInput(inputId = "y", 
                        label = "Y-axis:",
                        choices = c("suicides_no", "suicides_per_100k_pop"), 
                        selected = "suicides_per_100k_pop"),
            
            # Show data table
            checkboxInput(inputId = "show_data",
                          label = "Show Data: ",
                          value = FALSE)
            ),

        # Output
        mainPanel(
            
            # Show plot
            plotOutput(outputId = "plot"),
            
            # Show data table
            DT::dataTableOutput(outputId = "suicidetable")
        )
        
    )
)

# Define server function --------------------------------------------
server <- function(input, output) {
  
    # Create scatterplot object the plotOutput function is expecting
    output$plot <- renderPlot({
      
      if(input$x %in% c("Age", "Sex", "Generation")){
         ggplot(data = subset(suicides, country %in% input$country_data), 
                aes_string(x = input$x, y = input$y)) +
            geom_boxplot()
      }
      
      else{
         ggplot(data = subset(suicides, country %in% input$country_data), 
                aes_string(x = input$x, y = input$y)) +
            geom_smooth()
      }
    })
    
    # Print data table if checked
    output$suicidetable <- DT::renderDataTable(
        if(input$show_data){
            DT::datatable(data = subset(suicides, country %in% input$country_data), 
                          options = list(pageLength = 15), 
                          rownames = FALSE)
        }
    )
}

# Create the Shiny app object ---------------------------------------
shinyApp(ui, server)
```

### Conclusion

### Citations
https://bpr.berkeley.edu/2017/10/31/the-scourge-of-south-korea-stress-and-suicide-in-korean-society/
https://s3.amazonaws.com/happiness-report/2019/WHR19.pdf

